\documentclass{article}
\usepackage{mathnotes}

\title{Markov Chains}
\description{Stochastic processes with transition matrices, ergodic properties, and steady-state probability calculations using linear algebra methods.}

\begin{document}

\section{Markov Chains}

\begin{definition}[Markov Chain]
If we have a system with N states and for each state, a set of N probabilities from transitioning from that state to another state (including itself) we end up with an $N \times N$ matrix $P$ where the entry $P_{jk}, j,k \in \{1, \dots, N\}$ represents the probability of transitioning from state $j$ to state $k$.

Starting with some state, we can transition to another state, then another, with the probability of picking the next state dependent only on the existing state and the matrix $P.$ We call this sequence of random events a \textbf{Markov Chain}.

More formally, the \@{sequence} $X_1, X_2, \dots$ is called a Markov Chain if

\[ P(X_{n+1} = k|X_n = j, X_{n-1} = j_{n-1}, \dots, X_0 = j_0) = P(X_{n+1} = k | X_n = j). \]

We write

\[ p_{jk} = P(X_{n+1} = k | X_n = j), \]

and note that

\[ p_{jk} \geq 0 \]

and

\[ \sum_{k=1}^{N} p_{jk} =  1 \]

for $j = 1,2, \dots, N.$ We call the $p_{jk}$ values the transition probabilities of the Markov chain.
\end{definition}

\begin{note}
Note that $p_{jk}$ is the probability of going from state $j$ to state $k$ in one step. We use the notation

\[ p_{jk}^{(m)} \]

to represent the probability of going from state $j$ to state $k$ in $m$ steps.
\end{note}

\begin{theorem}
$p_{jk}^{(m)} = P_{jk}^m.$
\end{theorem}

\begin{proof}
Due to the magic of linear algebra, if we have a row vector $\vec{x}$ representing the probability of starting in each state, we can compute the probability of ending up in any given state after $m$ steps as

\[ \vec{x} \cdot P^m. \]
\end{proof}

\subsection{Ergodic Markov Chains}

\begin{definition}[Ergodic]
If for some positive integer $m$ we have that $p_{jk}^{(m)} > 0$ for all $j,k = 1,2,\dots,N,$ then the \@{Markov-chain} is said to be \textbf{ergodic.}
\end{definition}

\begin{theorem}[Stationary Distribution of an Ergodic Chain]\label{stationary-distribution}
When the \@{markov-chain} is \@{ergodic}, the \@[limit]{limit-sequence}

\[ \pi_k = \lim_{n \to \infty} p_{jk}^{(n)} \]

exists and

\[ \sum_{k=1}^N \pi_k = 1. \]
\end{theorem}

\begin{note}
Here, $\vec{\pi} = \langle \pi_1, \pi_2, \dots, \pi_N \rangle$ and the entries in it represent the long term probabilities of being in any given state, or equivalently, the portion of time spent in any given state.

Note that $\vec{\pi} \cdot P = \vec{\pi},$ which when combined with the fact that the entries of $\vec{\pi}$ sum to 1 lets us solve a linear system of equations to find the entries of $\vec{\pi}.$
\end{note}

\begin{definition}[Stochastic Matrix]\synonyms{"probability matrix", "transition matrix", "Markov matrix", "right stochastic matrix", "row stochastic matrix"}
A \textbf{stochastic matrix} (actually, a right stochastic matrix) is a \@{square-matrix} $A = (a_{ij}) \in \reals^{n \times n}$ with nonnegative \@{entries} whose rows each sum to one:

\[ a_{ij} \geq 0 \quad \text{for all } i,j, \qquad \sum_{j} a_{ij} = 1 \quad \text{for all } i. \]

Equivalently, $A\mathbf{1} = \mathbf{1}$ with $A \geq 0$ entrywise, where $\mathbf{1}$ is the all-ones vector.
\end{definition}

\begin{definition}[Left Stochastic Matrix]\synonyms{"column stochastic matrix"}
A \textbf{left stochastic matrix} is a \@{square-matrix} $A = (a_{ij}) \in \reals^{n \times n}$ with nonnegative \@{entries} whose columns each sum to one:
\[ a_{ij} \geq 0 \quad \text{for all } i,j, \qquad \sum_{i} a_{ij} = 1 \quad \text{for all } j. \]
Equivalently, $\mathbf{1}^\top A = \mathbf{1}^\top$ with $A \geq 0$ entrywise, where $\mathbf{1}$ is the all-ones vector.
\end{definition}

\begin{note}
A \@{stochastic-matrix} can be used to describe the transitions of a \@{markov-chain}.
\end{note}

\begin{definition}[Doubly Stochastic Matrix]\synonyms{"bistochastic matrix"}
If a \@{matrix} is both a \@{row-stochastic-matrix} and a \@{column-stochastic-matrix}, it is said to be a \textbf{doubly stochastic matrix.} That is, both its rows and its columns sum to 1.
\end{definition}

\begin{note}
A \@{doubly-stochastic-matrix} is a \@{convex-combination} of \@{permutation-matrices}.
\end{note}

A \@{stochastic-matrix} maps a probability vector to another probability vector — but only for the matching multiplication convention. A \@{row-stochastic-matrix} preserves the sum acting on the right, $\vec{x}^\top A$; a \@{column-stochastic-matrix} preserves it acting on the left, $A\vec{x}$; and a \@{doubly-stochastic-matrix} preserves it either way. The demo below lets you apply each type to a probability vector, build doubly-stochastic matrices as \@{convex-combinations} of permutation matrices, and iterate the map toward its stationary distribution.

\includedemo{stochastic-matrix}

\end{document}
